# Working with Data

In this section, we learn how to work with data in a **dataframe**. A dataframe is a two-dimensional array consisting of *rows* and *columns*. You can simply think of it as a spreadsheet (e.g. MS Excel, Google Sheets, etc.).

## Basic dataframes
R has some prebuilt functions to build dataframes. Let us see a simple example. Consider the following three vectors.

```{r}

name <- c("Sam", "Paulina", "Cenk")
age <- c(23, 34, 19)
height <- c(179, 167, 173)
```

Let us turn the data stored in different vectors into a single dataframe so that we can visualize the data better.

```{r}
#Let us first create the dataframe and assign it to the variable my_df
my_df <- data.frame(name,age,height)

#Let's print the dataframe now
my_df
```

We can select a particular row, column, or cell on a dataframe by using indices. For this we can use the slicing method `my_dataframe[row,column]`.


```{r}
#Let us select the entire first row 
my_df[1,]

```


```{r}
#Now, let us select the first column 
my_df[,1]

```

```{r}
#Now, let us find Paulina's height. For this, we need to get the 2nd row and 3rd column.
my_df[2,3]
```


```{r}
#Now, let us find Paulina's age and height. For this, we need to get the 2nd row and 2nd and 3rd columns.
my_df[2,2:3]
```

```{r}
#Finally, let us get Sam and Paulina's ages.
my_df[1:2,2]
```

You can also use the column name to select an entire column. Just add the dollar sign `$` after the df and then the column mane. 

```{r}
my_df$age
```



## Tibbles
The standard dataframes in R are good but not great. Often, we will deal with a lot of data we may not now which index to use to find the value we want. So, we need to be able to have some better ways to access data on our dataframes. We also want to be able to add new data or change some of the existing data easily. For this, we will use various packages in **tidyverse** for better dataframe management. 

Let us first load the tidyverse library, which will load the necessary packages for the functionality described in the following sections.


```{r, message=FALSE}
library(tidyverse)
```


Next, let us introduce tibbles. A  **tibble** is a dataframe with some improved properties. We can turn a regular dataframe into a tibble by calling the `as_tibble()` function on our dataframe.


```{r}
#Let's turn my_df into a tibble
my_tibble <- as_tibble(my_df)

#Let's print my_tibble
my_tibble
```

As you can see above, the console output tells you that this is a 3x3 tibble meaning that it has 3 rows and 3 columns. It also tells you the type of the data in each column. You can see the data types right under each column name. 


## Beyond Toy Data
So far we have been working with toy data. In real life projects, you will have a lot more data. The data will usually be stored in some file from which you will have to read into a dataframe. Alternatively, it might be some dataset that from a corpus easily accessible to R. Let us see a few ways in which we can load some realistic datasets into a tibble. 

### Reading data from a csv file

In this course, we will use some of the data sets from Bodo Winter's book. Go to [this website](https://osf.io/34mq9/) to download the `materials` folder. Once your data has been downloaded, navigate to the `materials/data` folder and locate the `nettle_1999_climate.csv` file.

To read in data from a csv to a tibble, we will use the `read_csv()` function. All we need to do is to provide the path to the csv file we want to read in. If your csv file is in the same folder as your script, you can simply give its name. Otherwise, you need to provide the relevant directory information as well in your path. 


```{r}

#Let's read in the data
nettle <- read_csv('/Users/umit/Desktop/materials/data/nettle_1999_climate.csv')

#Let's print the head of the data to see what it looks like
nettle
```

If you want to see the last 5 items, use the `tail()` function. 

```{r}
tail(nettle)
```


If you want to view the entire dataset, you can use `View(nettle)`. This will open a new tab in RStudio and show your data as a table. 


### Reading data from R data packages
R has various data packages you can install and use. Let us install the `languageR` which has some nice language datasets. Once you install the package and load the library, you can easily use the datasets as tibles. For all the details and available datasets in `languageR`, you can check the [languageR documentation](https://cran.r-project.org/web/packages/languageR/languageR.pdf) on CRAN.

```{r}
#Let's load the library
library(languageR)

#We'll use the dativeSimplified dataset, which is documented. Let's see the documentation
?dativeSimplified
```

```{r}
#let's use the dativeSimplified data from the languageR 
data <- as_tibble(dativeSimplified)

#Let's print the first few lines of the data
data
```

**Dative Alternation** is the phenomenon in English where a recipient of a ditransitive verb can occur as an NP or a PP.

1. Alex gave Sam a book.
2. Alex gave a book to Sam.

Both of these constructions are grammatical and they mean essentially the same thing. The question is what factors are involved in picking one of the forms over the other. Bresnan et al. (2007) used this data to determine the relevant factors. 
Let us randomly select 10 examples and see what they look like. For that, we can use the folloing code.


```{r}
# store all possible row indices in a vector
indices_all <- 1:nrow(data)

# set the random seed to make the results reproducible
set.seed(123)

# choose 10 such numbers at random without replacement
indices_random <- sample(indices_all, size = 10)

# use them to index the data frame to get the corresponding rows
data[indices_random,]
```


## Summarizing Data
Looking at the summary statistics of your data is always a good first step. Let's take a look at the percentage of NP realizations of the recipient by animacy of the theme.

```{r}
# First, let's take a look at the key dependet variable (NP or PP)

unique(data$RealizationOfRec)
```


```{r}
# now, let's compute the percentages (perc_NP) and the number of observations in each subset
data%>% 
  group_by(AnimacyOfRec) %>% 
  summarize(perc_NP = mean(RealizationOfRec == "NP"), 
                   N = n()
                  )
```

**What do the results say?**

* There are a total of 822 instances of animate recipients.
* 63% of the animate recipients are NPs.


## Working with dplyr
One of the packages in the `tidyverse` is `dplyr`. We use it to do various manipulations on the data frames. Check out the [dplyr cheatsheet](https://nyu-cdsc.github.io/learningr/assets/data-transformation.pdf) for further details. 

The `arrange` function will arrange your data in an ascending order. 

```{r}
arrange(data)
```

You can arrange the data based on a particular column. In that case, you need to provide the column name.

```{r}
arrange(data,LengthOfTheme)
```


```{r}
arrange(data[1:10,], LengthOfTheme)
```

If you want to arrange things in a descending order, then you need to put the `desc()` function around the relevant column.

```{r}
arrange(data, desc(LengthOfTheme))
```

Another useful function is the `select()` function which allows you to create new dataframes using only columns you want. 

```{r}
#Create the new dataframe using select
df <- select(data, Verb, LengthOfTheme)

#print the head
df
```

Another useful function is `sample_n()` which randomly samples some number of datapoints. 

```{r}
sample_n(data, 5)
```


Two other useful functions are `group_by()` and `ungroup()`. 

```{r}
#Let's group a small portion of the data by the realization of recipient
group_by(data[1:5], RealizationOfRec)
```

Now let us group the data by verbs.


```{r}
data_grouped_by_verb <- group_by(data,Verb)
```


An important but complex function is the `summarize()` function.

1. It divides a grouped data frame into subsets, with each subset corresponding to one value of the grouping variable (or a combination of values for several grouping variables).
2. It computes one or several values we specify on each such subset.
3. It creates a new data frame and puts everything together. The first column of this new data frame consists of levels of our grouping variable. In the following columns, the summarize() function prints the results of the computations we have specified.

Try to guess the result of the following code. What will you see as an output? What will be the name of the columns?

```{r}
# summarize several variables
summarize(data_grouped_by_verb, 
          prop_animate_rec = mean( AnimacyOfRec == "animate" ),
          prop_animate_theme = mean( AnimacyOfTheme == "animate" ),
          N = n()
          )
```

Try to interpret the output of the following code.

```{r}

# compute the averages
summarize(data_grouped_by_verb, 
          prop_anim = mean(AnimacyOfRec == "animate"),
          prop_inanim = 1-prop_anim,
          prop_v_recip_anim = ifelse(prop_anim > 0.5, "high", "low")
          )

```

The last line uses the function ifelse(condition, value1, value2), which, for each element of the condition vector returns the corresponding element of the value1 vector if the condition is true at that element, or an element of vector2 otherwise.

mutate() proceeds similarly to summarize() in dividing a grouped dataset into subsets, but instead of computing one or several values for each subset, it creates or modifies a column.

The main difference between mutate() and summarize() is the output. While mutate() modifies the original and returns a modified version of it, summarize() creates a brand new data frame with one row for every combination of the the grouping variable values.

A very simple application of mutate() is to simply create a new column. In this case, we don’t even need to group.

```{r}
# these two lines performs exactly the same action, 
# except the latter stores the result in df 
data$is_realization_NP <- (data$RealizationOfRec == "NP" ) 
df <- mutate(data, is_realization_NP = (RealizationOfRec == "NP") )

head(df, 2)

```


One final useful function is the `filter()` function. It allows you to find rows by particular values of a column. 

```{r}
filter(data, is_realization_NP == FALSE)
```


```{r}
filter(data, LengthOfTheme > 3.5)
```

## Pipes

### The problem

- The code below is really hard to read, even harder to maintain, and `dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme` is a terribly long variable name.
```{r, error=T}
dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme <-
      group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme)
df <- summarize(dativeSimplified_grouped_by_AnimacyOfRec_and_AnimacyOfTheme, 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- This alternative is also quite bad. To read this code, you need to know which bracket matches which other bracket.
```{r, error=T}
df <- summarize(group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme), 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- One nested function call may be OK. But try to read this.
```{r, error=T}
df <- dplyr::summarize(group_by(mutate(dativeSimplified, long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ), long_theme),
              perc_NP = mean(RealizationOfRec == "NP")
              )
```

- Or consider this expression (`sqrt` is the square root.)
```{r, error=T}
sqrt(divide_by(sum(divide_by(2,3), multiply_by(2,3)), sum(3,4)))
```

- Luckily, there a better way to write this expression.

### Pipes
- The problem is that we have too many levels of embedding.
- In natural language we avoid multiple embeddings of that sort by making shorter sentences, and using anaphors to refer to previous discourse.
- The packages **dplyr** and **magrittr** provide a limited version of such functionality, and we'll need to use **pipe** operators (`%>%` and `%<>%`) to link expressions with an 'anaphoric dependency'.
- Whenever you see `%>%`, you can think about it as the following: "Take whatever is on the left side, and use it in the function that is on the right side."
```{r, error=T}
library(dplyr)
library(magrittr)
# Typical notation. Read as "Divide 10 by 2."
divide_by(10, 2)

# Equivalent pipe notation. Read as "Take 10, and divide it by 2."
10 %>% divide_by(., 2)

# Equivalent pipe notation. Read as "Take 2, and divide 10 by it."
2 %>% divide_by(10, .)
```

- If the dot operator occurs in the first argument slot, it can be omitted. (R has pro-drop.)
```{r, error=T}
# pipe notation with omission of '.'
10 %>% divide_by(2)
```

- Let's see how it can resolve the mess below. (Repetition of previous example.)
```{r, error=T}
df <- mutate(group_by(dativeSimplified, AnimacyOfRec, AnimacyOfTheme), 
                  perc_NP = mean(RealizationOfRec == "NP") )
df
```

- And here is the much more readable version of this code:
```{r, error=T}
df <-  dativeSimplified %>%
          mutate(., long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ) %>% 
          group_by(., long_theme) %>% 
          dplyr::summarize(., perc_NP = mean(RealizationOfRec == "NP") )
```
- We don't actually need the dot:
```{r, error=T}
df <-  dativeSimplified %>%
          mutate(long_theme = ifelse(LengthOfTheme > 1.6, "long", "short") ) %>% 
          group_by(long_theme) %>% 
          dplyr::summarize(perc_NP = mean(RealizationOfRec == "NP") )
```


- The `%<>%` operator is a convenient combination of `%>%` and `<-` which you can use to directly modify an object.

```{r, error=T}
# load the package magrittr in order to access the assignment pipe operator
library(magrittr)

# create a vector with numbers from 1 to 10
x <- 1:10
# keep only numbers < 5:
#   (i) without %<>%
x <- x[x<5]
#   (i) with %<>%
x %<>% .[.<5]
```

```{r, error=T}
# lets add several columns to 'dativeSimplified'
dativeSimplified %<>% mutate(A=1, B=2, C=3, D=4)
head(dativeSimplified)
```