# Plotting
(`r format(Sys.time(), '%d %B, %Y, %H:%M')`)

```{css echo=FALSE, include}
.r-output {
  font-weight: bold;
}
.function {
  background-color: lightgrey;
  padding: 0.5em 0.5em 0.5em 4em;
  border: 1px solid black;
}
```



## The basics of ggplot2

- Let's first take a look at some example plots.
- Create a synthetic data set and load the `ggplot2` package to access the plotting functionality.
```{r, error=T}
library(ggplot2)
df <- data.frame(a=1:10, a_sq=(1:10)^2, my_group = c("weekday","weekend"))
df
```

- Take a look at the following code and the resulting plots. Can you tell what parts that start with geom_... does?

```{r, error=T}
ggplot(data = df, mapping = aes(x = a, y = a_sq)) + geom_point()
```
```{r, error=T}
ggplot(data = df, mapping = aes(x = a, y = a_sq)) + geom_bar(stat="identity")
```
```{r, error=T}
ggplot(data = df, mapping = aes(x = a_sq, y = a)) + geom_point()
```
```{r, error=T}
ggplot(data = df, mapping = aes(x = a_sq, y = a)) + geom_bar(stat="identity")
```

## The basics of ggplot2
- So what do those function calls mean?

- Let's take a look at it again: This is pretty much the minimal useful plotting command in R.
```{r, eval=F, error=T}
ggplot(data = df, mapping = aes(x = a, y = a_sq)) + geom_point()
```

- Each ggplot2 plot specification consists, at a minimum, of three parts:
  1. <span style="color:blue">the data to plot</span>
  2. <span style="color:red">*an abstract specification of the plot* (a rough mapping between variables and axes and other plot elements, such as *groups*, *facets*, etc.)</span>
  3. <span style="color:green">*concrete instructions on what to draw* (a specification of the actual visual elements to use)</span>

- They correspond to three parts of the `ggplot()` function call 
  1. <span style="color:blue">**data:** `data = df`</span>
  2. <span style="color:red">**'aesthetic':** `mapping = aes(x, y)`</span> 
  3. <span style="color:green">**'geoms':** `+ geom_point()`</span>

- You can read the instruction below as *"Create a plot <span style="color:blue">using the data in data frame df</span>, <span style="color:red">placing `a` on the x-axis and `a_sq` on the y-axis</span>, <span style="color:green">and visualize the data using points</span>"*.
- Keep in mind that information regarding x and y axes is specified within a function called `aes()`. 

```{r, eval=F, error=T}
ggplot(data = df, mapping = aes(x = a, y = a_sq)) + geom_point()
```
- As an aside: A shorter way to write the same code is below, and I'll mostly use some mixed form.
```{r, eval=F, error=T}
ggplot(df, aes(a, a_sq)) + geom_point()
```


## Using lines in plots

- We already know `geom_point` and `geom_bar`. Let's take a look at some other *geoms*:.

- `geom_line` connects the (invisible, in this case) points in the plot.
```{r, error=T}
ggplot(df, aes(a, a_sq)) + geom_line()
```

- We can even combine geoms:
```{r, error=T}
ggplot(df, aes(a, a_sq)) + geom_point() + geom_line()
```

- ... in fact, as many as we want. But there is no guarantee that the result will look good, or even make sense.
```{r, error=T}
ggplot(df, aes(a, a_sq)) + geom_point() + geom_line() + geom_bar(stat = "identity")
```

- The order of their specification matter a little bit. Here, the line is plotted over the bars, in contrast to the previous plot.
```{r, error=T}
ggplot(df, aes(a, a_sq)) + geom_point() + geom_bar(stat = "identity") + geom_line()
```


## Color and fill
- Relationships between two variables are usually easy to visualize, but often there is a third variable.
- There are various ways for dealing with it.

- Let's first try using color coding for the third variable. 
```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_point(stat = "identity")
```

- Let's try this with bar plots. Not at all what you expected, is it? 

```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_bar(stat = "identity")
```

- This is what we wanted. The right argument for bar plots is `fill`.
```{r, error=T}
ggplot(df, aes(a, a_sq, fill = my_group)) + geom_bar(stat = "identity")
```


- So why isn't the aesthetic argument for bar plots not also `color`?
- Because geoms in ggplot2 have `fill` (the color of the inner part of the object), and a `color` (the color of the line with which they are drawn).

- Points don't have a fill. (Don't ask me why.)
- We can try, if you do not believe me. See that even though we specify a `fill` argument for geom_point, `color` argument overwrites it. 
 

```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_point(size=10, fill = "black")
```
- If points had a fill, we would expect the argument that comes last to overwrite the previous one. 
- Bars have both fill and color arguments. 

```{r, error=T}
ggplot(df, aes(a, a_sq, fill = my_group)) + geom_bar(stat="identity", color = "black")
```
```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_bar(stat="identity", fill = "black")
```

## Grouping and facets

- Color, fill, etc. implicitly group the data set into different subgroups.
- You can see that better if you connect the points by lines.
```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_point()  + geom_line()
```

- This can be done explicitly as well.
```{r, error=T}
ggplot(df, aes(a, a_sq, group = my_group)) + geom_point()  + geom_line()
```

- Now it's very hard to see which line is which, so let's at least separate it into different **facets** (aka *'panels'*). 
- We can introduce our new facets with the function `facet_wrap()`. Keep in mind that the grouping variable is introduced with `~`.
- The name of the groups can be seen at the top of the plots. 

```{r, error=T}
ggplot(df, aes(a, a_sq, color = my_group)) + geom_point()  + geom_line() + facet_wrap(~my_group)
```