--- 
title: "Ling 411 - Fall 2022"
author: "Ümit Atlamaz"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
description: |
  This is an R tutorial for the Ling 411 Class at Boğaziçi University in Fall 2022.
link-citations: yes
github-repo: rstudio/ling_411
---

```{r z, include = T}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, class.output="r-output")
source("./source/r_functions.R")
```



# Getting Started

```{css echo=FALSE}
.r-output {
  font-weight: bold;
}
.function {
  background-color: lightgrey;
  padding: 0.5em 0.5em 0.5em 4em;
  border: 1px solid black;
}
```

Welcome to the R tutorial for Ling 411. The purpose of these lecture notes is to help remind you some of the R related material we covered in the class. The material here is not intended to be complete and self-contained. These are just lecture notes. You need to attend the classes and Problem Sessions to get a full grasp of the concepts.


## Some great resources
- Throughout the semester, I will draw on from the following resources. These are just useful resources and feel free to take a look at them as you wish.
  - The great introduction materials developed at the University of Glasgow: <https://psyteachr.github.io/>, in particular ['Data Skills for Reproducible Science'](https://psyteachr.github.io/msc-data-skills/).
  - The also pretty great introduction to R and statistics by Danielle Navarro available [here](https://learningstatisticswithr.com).
  - Matt Crump's ['Answering Questions with Data'](https://crumplab.github.io/statistics/).
  - Primers on a variety of topics: https://rstudio.cloud/learn/primers
  - Cheat sheets on a variety of topics: https://rstudio.cloud/learn/cheat-sheets

- The following tutorials are great too.
  - ['The Tidyverse Cookbook'](https://rstudio-education.github.io/tidyverse-cookbook/)
  - ['A Ggplot2 Tutorial for Beautiful Plotting in R'](https://cedricscherer.netlify.app/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/)
  - ['R Graphics Cookbook, 2nd edition'](https://r-graphics.org/)
  
  
## Blocks
Code, output, and special functions will be shown in designated boxes. The first box below illustrates a **code block**. The code block contains code that you can type in your R interpreter as the source code. You can simply copy and paste it in your R code. The second box indicates the **output** of R given the code in the first box.

```{r addition}
2+2
```

Functions will be introduced in grey boxes. The following grey box describes the `summary()` function.

`r function_definition("summary(x)", "Returns the summary statistics of a dataframe.", c(x="A dataframe."))`
\
\
The following code block uses the `summary()` function on the `mtcars` dataframe that comes pre-installed with R. 


```{r cars1}
summary(mtcars)
```
\
If you want to learn more about the `mtcars` dataset, you can simply put a question mark in front of its name, which will show the documentation for the dataset. The documentation will pop up in the `Help` tab on the bottom right window in RStudio.

```{r mtcars2}
?mtcars
```


